---
title: "Classifying inbred mice into sustained and phasic anxiety phenotypes using anxiotraitR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial demonstrates how to perform behavioral classification of female and male mice according to their trait anxiety  using the R package anxiotraitR. The experimental pipeline to obtain the behavioral data necessary for classification is described in detail in our publication Kovlyagina et al. [1]. Briefly, we measure individual freezing responses of mice in memory retrieval sessions with a prolonged conditioned stimulus (CS) exposure of 6 min following auditory aversive conditioning (AAC). Subsequently, we employ log-linear regression to estimate the intercept and slope (regression coefficient) of the fitted freezing curve during CS exposure. Together with the average freezing during the last 3.5 min of CS exposure, we use these parameters to classify mice into sustained and phasic responders using our pre-trained machine learning (ML) models implemented in anxiotraitR [2].

## Installation

To install anxiotraitR, devtools must be available. The following code sequence will check for a devtools installation and then will install anxiotraitR directly from GitHub [2]:

```
if(!require(devtools)){
    install.packages("devtools")
    library(devtools)
}
devtools::install_github("johannesmiedema/anxiotraitR")
```

## Classification 
To classify mice into sustained and phasic responders, a data frame must be provided, where each row represents the freezing values of individual animals during CS presentation. To demonstrate this, we have included an example dataset. As seen below, the data frame contains 36 columns, each corresponding to a freezing value as a percentage in 30 s time bins during MR1 AAC.

$~$

```
library(anxiotraitR)

mydata <- anxiotraitR::freezingDataset
head(mydata)
```

```{r, layout="l-body-outset", echo=FALSE}
suppressWarnings(library(knitr))
library(anxiotraitR)
mydata <- anxiotraitR::freezingDataset
head(mydata)
```

$~$

Classification of animals into sustained and phasic responders is performed only using freezing responses measured during CS presentation. In our example, CS occurred in bins 13 to 24. Therefore, the data frame must be subset to contain only columns 13:24 (please note that the data frame may need to be subset differently depending on the CS bins in the respective experiment) :
 
$~$
 
 
```
mydata_subsetted <- mydata[,13:24]
head(mydata_subsetted)
```

```{r, layout="l-body-outset", echo=FALSE}
suppressWarnings(library(knitr))
library(anxiotraitR)
mydata <- anxiotraitR::freezingDataset
mydata_subsetted <- mydata[,13:24]
head(mydata_subsetted)
```


$~$

This data frame can now be used for classification. The example dataset was obtained from female animals during MR 1. These parameters need to be specified to select the correct pre-trained model for classification:

$~$

```
myclassification <- anxiotraitR::classify_freezer(data_MR1 = mydata_subsetted, sex = "female", MR = 1)
head(myclassification)
```
$~$

The resulting object contains the inferred phenotype of each animal as a vector in the same order as provided by the rows in the input data frame. Furthermore, the parameters needed for the machine learning model including the intercept and slope of the fitted freezing curve and average freezing during the last 3 min of CS presentation are included in the output of the classification function.

Classification results can be visualized with a bivariate plot of average freezing vs. fitted freezing curve slope (to which we refer as decay rate of freezing) or as actual freezing curves over time bins for sustained and phasic responders.
For classification with male animals or different MR sessions (MR1 or MR2), the parameters of the function must be specified accordingly. For MR2 classification, the data frame needs to be specified under the argument data_MR2:

For classification with male animals or different MR sessions (MR1 or MR2), the parameters of the function must be specified accordingly. For MR2 classification, the data frame needs to be specified under the argument data_MR2:

$~$

```
myclassification_MR2 <- PhenoFreeze::classify_freezer(data_MR2 = mydata_subsetted, sex = "female", MR = 2)

```
$~$

## Retraining PhenoFreeze

Apart from classification with pre-trained models, it is also possible to retrain the existing models by adding new data. To use the retrain_model function, the user must have ggplot2 installed and loaded, as this function will also produce plots for evaluation of the trained models. Due to the small sizes of the data sets, Monte-Carlo Cross-Validation is used to assess the performance of the used models. Per default, the number of iterations is 1000, but can be changed with the training_iterations parameter within the function call. As an example in this tutorial, the loaded dataset from female animals will be added to the male MR1 training data for demonstration purposes:

$~$

```
#ggplot2 needs to be loaded
library(ggplot2)


#Retrain male MR1 models by using the data obtained from the example dataset
retrainingresults <- PhenoFreeze::retrain_model(new_data = mydata_subsetted, sex = "male", MR = 1)

```
$~$

The resulting object can be used to assess the metric plots, metric tables and the trained models themself. For example, to investigate the plot for the accuracy metric:

$~$


```
#Obtain Accuracy plot
Accuracy <- retrainingresults$Accuracy

#Show the Accuracy plot
Accuracy

```
$~$

This can be done accordingly for Specificity, Sensitivity, F1 and Auroc. The plots can be accessed accordingly with the dollar operator of the resulting object of the retrain_model() function. The metrics can also be viewed as a dataframe which shows the mean and +/- standard deviation. This can be done accordingly:

$~$

```
#Obtain model performance metrics as a dataframe
Stats <- retrainingresults$Stats
Stats

```

After investigating all metrics, the user can save the best-performing model as an RDS object. First, the model needs to assigned to a new variable. For example, we will save the radial SVM model (RSVM) in this tutorial. The other models can be accessed via the dollar operator by specifing "SVM" for the SVM with a linear kernel, "RF" for random forest, "Logistic" for the logistic regression model, and "LDA" for linear discriminant analysis.

$~$

```
mymodel <- retrainingresults$RSVM

#Save as RDS object: the second parameter specifies the name of the saved object and the path 
saveRDS(mymodel, "mymodel.RDS")

```
$~$

After saving the retrained model as an RDS object, this model can now be used for custom classification using the classify_freezer() function. For this, the model name and path of the saved RDS object needs to be specified, as well as the data for custom classification, which needs to be specified under the data variable. The other parameters shown before, like sex and MR are not needed in this step:

$~$


```
newclassification <- PhenoFreeze::classify_freezer(model = "mymodel.RDS", data = mydata_subsetted)

```

## References
